Summary of the things that I learned

### 2024-11-25 100-introduction.tex
Sort out simeq and cong
Minor changes, make the statement about the Karoubi envelope more subtle

However, because the computation part of such a proof involves a lot of code, and computer code tends to contain bugs, accepting a proof by computer involves a certain amount of trust. Therefore, both of the theorems mentioned have subsequently also been proved using `computer proof assistants' \autocite{formalized-four-color-theorem, formalized-Kepler-conjecture}. Such proof assistants, like `rocq', `lean' and `Agda', are computer programs with only a very small `trusted codebase'\footnote{The trusted codebase is the part of the codebase of which we have to trust that it is correct. The rest of the codebase is checked using this trusted core, so if the trusted core is correct, the rest is too.}, that can verify mathematical reasoning. In this way, if we trust the small core of such a proof assistant, and the proof assistant says that a proof is correct, then we can trust that indeed, the proof is correct.

@@

Now, much of contemporary mathematics is built on the foundation of set theory, which we will often refer to as \iindex{classical mathematics}. Usually, `set theory' refers to ZFC: Zermelo Fraenkel set theory with the axiom of choice, developed in the early 1900s \autocite{Zermelo-set-theory}. On the other hand, most of the formalization in proof assistants is built on the foundation of type theory. This is because computers can reason better with types than with sets, as evidenced by the many typed programming languages that are around nowadays. However, type theory has been around longer than computers have: it was initially developed by Bertrand Russell in the early 1900s \autocite[][Appendix B]{Russell-type-theory}, to create a foundation of mathematics that avoided Russell's paradox\footnote{Russell's paradox is the question ``Consider the set of all sets that do not contain themself. Does this set contain itself?'' Answering this question with either `yes' or `no' leads to a contradiction.}.

@@

In this thesis, we study the paper `Classical lambda calculus in modern dress' through the lens of univalent foundations, work out the details of the proofs, and formalize part of the paper and its preliminaries. The major contributions are:
\begin{itemize}
  \item More detailed versions of Hyland's definitions and proofs, complemented with some examples (Chapters \ref{ch:algebraic-structures} and \ref{ch:the-paper}). Also, the work of Dana Scott and Paul Taylor that Hyland expands upon (Chapter \ref{ch:previous-work}), and most of the category theoretical preliminaries needed to understand the paper (Chapter \ref{ch:category-theory}), to make this thesis reasonably self-contained. All this makes Hyland's paper more accessible to computer scientists.
  \item The translation of Hyland's paper and the work of his predecessors to univalent foundations (Chapter \ref{ch:the-paper}). This shows that his work can be translated to univalent foundations, and contributes to the knowledge about translating classical mathematics to univalent foundations in general.
  \item A formalization of part of the paper in rocq (Chapter \ref{ch:the-formalization}). This contributes to the knowledge about formalization in general.
  \item A new proof for the fundamental theorem of the $ \lambda $-calculus (Section \ref{sec:elementary-fundamental-theorem}). This version is easier to read and verify, because it is more elementary and uses less category theory than Hyland's proof.
  \item An analysis of the behaviour of the Karoubi envelope in univalent foundations in general (Section \ref{sec:karoubi-envelope}), and in the specific case of Paul Taylor's and Martin Hyland's work (Remark \ref{rem:difference-Taylor-Hyland}). Here we see an example of two definitions that are equivalent in classical mathematics, but become different in univalent foundations. It is interesting to see the differences between the behaviour of the two definitions in univalent foundations.
\end{itemize}

>>>>

However, because the computation part of such a proof involves a lot of code, and computer code tends to contain bugs, accepting a proof by computer involves a certain amount of trust. Therefore, both of the theorems mentioned have subsequently also been proved using `computer proof assistants' \autocite{formalized-four-color-theorem, formalized-Kepler-conjecture}. Such proof assistants, like `rocq', `lean' and `Agda', are computer programs with only a very small `trusted codebase'\footnote{The trusted core is the part of the codebase of which we have to trust that it is correct. Among other things, it verifies the output of every other part of the codebase, so if the trusted core is correct, the rest is too.}, that can verify mathematical reasoning. In this way, if we trust the small core of such a proof assistant, and the proof assistant says that a proof is correct, then we can trust that indeed, the proof is correct.

@@

Now, much of contemporary mathematics is built on the foundation of set theory with classical logic, which we will often refer to as \iindex{classical mathematics}. Usually, `set theory' refers to ZFC: Zermelo Fraenkel set theory with the axiom of choice, developed in the early 1900s \autocite{Zermelo-set-theory}. On the other hand, most of the formalization in proof assistants is built on the foundation of type theory. This is because computers can reason better with types than with sets, as evidenced by the many typed programming languages that are around nowadays. However, type theory has been around longer than computers have: it was initially developed by Bertrand Russell in the early 1900s \autocite[][Appendix B]{Russell-type-theory}, to create a foundation of mathematics that avoided Russell's paradox\footnote{Russell's paradox is the question ``Consider the set of all sets that do not contain themself. Does this set contain itself?'' Answering this question with either `yes' or `no' leads to a contradiction.}.

@@

In this thesis, we study the paper `Classical lambda calculus in modern dress' through the lens of univalent foundations, work out the details of the proofs, and formalize part of the paper and its preliminaries. The major contributions are:
\begin{itemize}
  \item More detailed versions of Hyland's definitions and proofs, complemented with some examples (Chapters \ref{ch:algebraic-structures} and \ref{ch:the-paper}). Also, the work of Dana Scott and Paul Taylor that Hyland expands upon (Chapter \ref{ch:previous-work}), and most of the category theoretical preliminaries needed to understand the paper (Chapter \ref{ch:category-theory}), to make this thesis reasonably self-contained. All this makes Hyland's paper more accessible to computer scientists.
  \item The translation of Hyland's paper and the work of his predecessors to univalent foundations (Chapter \ref{ch:the-paper}). This shows that his work can be translated to univalent foundations, and contributes to the knowledge about translating classical mathematics to univalent foundations in general.
  \item A formalization of part of the paper in rocq (Chapter \ref{ch:the-formalization}). Of particular interest here is a tactic for applying $ \beta $-reduction and substitution to $ \lambda $-terms (Section \ref{sec:tactic}). This formalization contributes to the knowledge about formalization in general.
  \item A new proof for the fundamental theorem of the $ \lambda $-calculus (Section \ref{sec:elementary-fundamental-theorem}). This version is easier to read and verify, because it is more elementary and uses less category theory than Hyland's proof.
  \item An analysis of the behaviour of the Karoubi envelope in univalent foundations in general (Section \ref{sec:karoubi-envelope}), and in the specific case of Paul Taylor's and Martin Hyland's work (Remark \ref{rem:difference-Taylor-Hyland}). It is one of the places where subtleties arise when translating from classical mathematics to univalent foundations.
\end{itemize}

### 2024-11-25 200-category-theory.tex
Minor changes, make the statement about the Karoubi envelope more subtle

\subsection{Weak Equivalences}
There is also the notion of `weak equivalence'. In some cases, this is equivalent to an adjoint equivalence (for example, when its domain is univalent, see Section \ref{sec:univalence-principle}).
\begin{definition}
  A functor $ F: \C \to \D $ is called a \iindex{weak equivalence} if it is essentially surjective and fully faithful.
\end{definition}

@@

Let $ \A $, $ \B $ and $ \C $ be categories and let $ F : \A \to \B $ be a functor.
\begin{definition}
  Precomposition gives a functor between functor categories $ F_* : [\B, \C] \to [\A, \C] $. If $ F_* $ has a left adjoint, we will denote call this adjoint functor the \textit{left Kan extension}\index{Kan extension!left} along $ F $ and denote it $ \mathrm{Lan}_F : [\A, \C] \to [\B, \C] $.

@@

\begin{remark}\label{rem:karoubi-classical}
  The situation in classical mathematics is very similar to the situation above: We can define a category $ \bar \D $ which is defined using idempotents, a category $ \tilde \D $ which consists of presheaves with chosen retractions, and a category $ \hat \D $ which is the image of $ \tilde \D $ under the forgetful functor $ F : \hat \D \to P \C $. More specifically, $ \hat \D $ is the full subcategory of $ P \C $, which we get if we only select the presheaves $ F $ such that we can find some $ X : \C $ and some retraction $ (r, s) : F \triangleleft \yo(X) $. We again have an adjoint equivalence $ \bar \D \simeq \tilde \D $ and a fully faithful, essentially surjective functor $ F : \tilde \D \to \hat \D $.

@@

For convenience, the lemma below works very abstractly with pointwise left Kan extension using colimits, but one could also prove this using just (co)equalizers (or right Kan extension using limits).
\begin{lemma}[\coqident{AlgebraicTheories.FundamentalTheorem.CommonUtilities.KaroubiEnvelope}{karoubi_pullback_equivalence}]
  Let $ \D $ be a category with colimits. We have an adjoint equivalence between $ [\C, \D] $ and $ [\overline \C, \D] $.
\end{lemma}
\begin{proof}
  We already have an adjunction $ \Lan {\iota_\C} {} \dashv \iota_{\C*} $. Also, since $ \iota_\C $ is fully faithful, we know that $ \eta $ is a natural isomorphism. Therefore, we only have to show that $ \epsilon $ is a natural isomorphism. That is, we need to show that $ \epsilon_G(X, a): \D(\Lan {\iota_\C} (\iota_{\C*} G) (X, a), G(X, a)) $ is an isomorphism for all $ G: [\overline \C, \D] $ and $ (X, a): \overline \C $.

>>>>

\subsection{Weak Equivalences}
There is also the notion of `weak equivalence'. In some cases, this is equivalent to an adjoint equivalence (for example, when its domain is univalent, see Section \ref{sec:univalence-principle}).
\begin{definition}
  A functor $ F: \C \to \D $ is called a \iindex{weak equivalence} if it is essentially surjective and fully faithful. We will sometimes denote a weak equivalence with
  \begin{center}
    \begin{tikzcd}
      \C \arrow[r, hook, two heads] & \D
    \end{tikzcd}
  \end{center}
  because intuitively it is injective and surjective, up to isomorphism.
\end{definition}

@@

Let $ \A $, $ \B $ and $ \C $ be categories and let $ F : \A \to \B $ be a functor.
\begin{definition}
  Precomposition gives a functor between functor categories $ F_* : [\B, \C] \to [\A, \C] $. If $ F_* $ has a left adjoint, we will call this adjoint functor the \textit{left Kan extension}\index{Kan extension!left} along $ F $ and denote it $ \mathrm{Lan}_F : [\A, \C] \to [\B, \C] $.

@@

\begin{remark}\label{rem:karoubi-classical}
  The situation in classical mathematics is very similar to the situation above: We can define a category $ \bar \D $ which is defined using idempotents, a category $ \tilde \D $ which consists of presheaves with chosen retractions, and a category $ \hat \D $ which is the image of $ \tilde \D $ under the forgetful functor $ F : \hat \D \to P \C $. To be more precise, $ \hat \D $ is the full subcategory of $ P \C $, which we get if we only select the presheaves $ F $ such that we can find some $ X : \C $ and some retraction $ (r, s) : F \triangleleft \yo(X) $. We again have an adjoint equivalence $ \bar \D \simeq \tilde \D $ and a fully faithful, essentially surjective functor $ F : \tilde \D \to \hat \D $.

@@

For convenience, the lemma below works very abstractly with pointwise left Kan extension using colimits, but one could also prove this using just (co)equalizers (or right Kan extension using $ \C $-small limits).
\begin{lemma}[\coqident{AlgebraicTheories.FundamentalTheorem.CommonUtilities.KaroubiEnvelope}{karoubi_pullback_equivalence}]
  Let $ \D $ be a category with $ \C $-small colimits. We have an adjoint equivalence between $ [\C, \D] $ and $ [\overline \C, \D] $.
\end{lemma}
\begin{proof}
  We already have an adjunction $ \Lan {\iota_\C} {} \dashv \iota_{\C*} $. Also, since $ \iota_\C $ is fully faithful, we know that $ \eta $ is a natural isomorphism. Therefore, we only have to show that $ \epsilon $ is a natural isomorphism. That is, we need to show that $ \epsilon_G(X, a): \D(\Lan {\iota_\C} (\iota_{\C*} G) (X, a), G(X, a)) $ is an isomorphism for all $ G: [\overline \C, \D] $ and $ (X, a): \overline \C $.

### 2024-11-25 500-previous-work.tex
Minor changes, make the statement about the Karoubi envelope more subtle

Types in the $ \lambda $-calculus correspond to objects in the category.

Types $ A \to B $ in the $ \lambda $-calculus correspond to exponential objects $ B^A $ in the category.

Terms in the $ \lambda $-calculus of type $ B $, with free variables $ x_1: A_1, \dots, x_n: A_n $, correspond to morphisms $ A_1 \times \dots \times A_n \to B $.

A free variable $ x_i: A_i $ in a context with free variables $ x_1: A_1, \dots, x_n: A_n $ corresponds to the projection morphism $ \pi_i : A_1 \times \dots \times A_n \to A_i $.

Given a term $ s: B_1 \to B_2 $ and a term $ t: B_1 $, both with free variables $ x_1: A_1, \dots, x_n: A_n $, corresponding to morphisms $ \overline s: A_1 \times \dots \times A_n \to B_2 $ and $ \overline t: A_1 \times \dots \times A_n \to B_1 $, the application $ st: B_2 $ corresponds to the composite of the product morphism with the evaluation morphism $ A_1 \times \dots \times A_n \to B_2^{B_1} \times B_1 \to B_2 $.
\begin{center}
  \begin{tikzcd}
    & A_1 \times \dots \times A_n \arrow[ld, "\overline s"'] \arrow[rd, "\overline t"] \arrow[d, "{\langle \overline s, \overline t \rangle}", dashed] &\\
    B_2^{B_1} & B_2^{B_1} \times B_1 \arrow[l, "\pi_1"] \arrow[r, "\pi_2"'] \arrow[d, "ev"] & B_1\\
    & B_2 &
  \end{tikzcd}
\end{center}

Given a term $ t: B $ with free variables $ x_1: A_1, \dots, x_n: A_n $, the abstraction $ (\lambda x_n, t): A_n \to B $ corresponds to using the adjunction corresponding to the exponential object of $ A_n $:
\[ C(A_1 \times \dots \times A_{n-1} \times A_n, B) \cong C(A_1 \times \dots \times A_{n-1}, B^{A_n}). \]

@@

\begin{definition}[\coqident{AlgebraicTheories.CategoryOfRetracts}{R}]
  Note that the set of $ \lambda $-terms without free variables $ L_0 $ has a monoid structure under the composition defined above. The category \index{R}$ \R $ is defined as the Karoubi envelope $ \overline C_{L_0} $ of the category $ C_{L_0} $ of this monoid.

>>>>

\begin{itemize}
  \item Types in the $ \lambda $-calculus correspond to objects in the category.
  \item Types $ A \to B $ in the $ \lambda $-calculus correspond to exponential objects $ B^A $ in the category.
  \item Terms in the $ \lambda $-calculus of type $ B $, with free variables $ x_1: A_1, \dots, x_n: A_n $, correspond to morphisms $ A_1 \times \dots \times A_n \to B $.
  \item A free variable $ x_i: A_i $ in a context with free variables $ x_1: A_1, \dots, x_n: A_n $ corresponds to the projection morphism $ \pi_i : A_1 \times \dots \times A_n \to A_i $.
  \item Given a term $ s: B_1 \to B_2 $ and a term $ t: B_1 $, both with free variables $ x_1: A_1, \dots, x_n: A_n $, corresponding to morphisms $ \overline s: A_1 \times \dots \times A_n \to B_2^{B_1} $ and $ \overline t: A_1 \times \dots \times A_n \to B_1 $, the application $ st: B_2 $ corresponds to the composite of the product morphism with the evaluation morphism $ A_1 \times \dots \times A_n \to B_2^{B_1} \times B_1 \to B_2 $.
    \begin{center}
      \begin{tikzcd}
        & A_1 \times \dots \times A_n \arrow[ld, "\overline s"'] \arrow[rd, "\overline t"] \arrow[d, "{\langle \overline s, \overline t \rangle}", dashed] &\\
        B_2^{B_1} & B_2^{B_1} \times B_1 \arrow[l, "\pi_1"] \arrow[r, "\pi_2"'] \arrow[d, "ev"] & B_1\\
        & B_2 &
      \end{tikzcd}
    \end{center}
  \item Given a term $ t: B $ with free variables $ x_1: A_1, \dots, x_n: A_n $, the abstraction $ (\lambda x_n, t): A_n \to B $ corresponds to using the adjunction corresponding to the exponential object of $ A_n $:
    \[ \C(A_1 \times \dots \times A_{n-1} \times A_n, B) \cong \C(A_1 \times \dots \times A_{n-1}, B^{A_n}). \]
\end{itemize}

@@

\begin{definition}[\coqident{AlgebraicTheories.CategoryOfRetracts}{R}]
  Note that the set of $ \lambda $-terms without free variables $ L_0 $ has a monoid structure under the composition defined above. The category \index{R}$ \R $ is defined as the Karoubi envelope $ \overline C_{L_0} $ (Definition \ref{def:karoubi}) of the category $ C_{L_0} $ of this monoid.

### 2024-11-25 700-formalization.tex
Minor changes, make the statement about the Karoubi envelope more subtle

\subsection{Propagation of Substitution}
Once the pure $ \lambda $-calculus is defined, it is not very hard to give it a $ \lambda $-theory structure, using the induction principle a couple of times to show that some identities are satisfied.

>>>>

\subsection{Propagation of Substitution}\label{sec:tactic}
Once the pure $ \lambda $-calculus is defined, it is not very hard to give it a $ \lambda $-theory structure, using the induction principle a couple of times to show that some identities are satisfied.
